Acceleration of bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-11 trunk/LateX/paper8.tex
latest draft after changes from Dr. Prasad

\begin{document}
\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures. bilateral filtering kernel is used to smooth an image without blurring its edges. It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.

@@

In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half. We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions. We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:
\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

@@

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

@@

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.

@@

It can be seen that after taking the $exp$ instruction out of the computation, the total sequential execution time is consistently less than 2\% of the original execution time with $exp$ instruction in place. For the smaller image it is strictly less than 1\% of the original time with $exp$ instruction in place. To be precise it is 1.39\% maximum on Intel Nehalem for the larger image and 0.65\% for the smaller image on Core i7 processor. The lower memory transfer time for smaller image is due to the fact that the smaller image fits into the cache memory entirely and hence there are no repeated data transfers between the processor and the main memory.

@@

We have implemented a number of variations of bilateral filtering kernel\footnote[2]{Our bilateral filtering code is open-source licensed under GPL and available from online repository \url{http://code.google.com/p/blfilter}.}  for multicore chips, results of which are reported in section \ref{sec:results}. These algorithms are labeled as follows: 
\begin{itemize}
\item $BL\_Naive$ is the sequential version of the naive bilateral filtering algorithm, as described by Tomasi\cite{Tomasi1998}. 
\item $BL\_Tiles$ is an improved version of bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs. 
\item $BL\_Assembly$ is the version of naive bilateral filtering kernel implemented using assembly level code. 
\item $BL\_A\_Reduction$ is the version of bilateral filtering kernel that uses assembly language as well as reduction optimizations discussed in Section~\ref{sub:reduction}.
\item $BL\_Intrinsic$ version is an extension of $BL\_Assembly$ version where SSE intrinsics have replaced assembly level instructions.
\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that employs intrinsics for reduction methods. These intrinsics can compute the sum of two 128 bit operands in parallel by exploiting the vector array processing. 
\end{itemize}
$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
 $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
After moving the redundant calculation out of the loop the same instruction becomes
 $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$

\subsubsection{Input Image}
In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images. One that fits into cache easily and the other that does not fit into cache but fits into main memory. Both of the images were converted to single spectrum images represented by single precision numbers for each pixel. The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it. As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB. The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it. None of the multicore cache memories were big enough to store this image entirely.
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Results}
\label{sec:results}
\subsection{Performance of bilateral filtering kernel on CPU}
For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21. The spatial spread and the photometric spread were also set to 10. The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization. For Core i7 chip compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading, i.e. there are no more than one thread per core.

@@

Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., achieving timings no less than 75\% of the naive algorithm's timings.

To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code - CUDA's quasi-assembly language - of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

Another challenge in interpreting our results was porting the concept of \lq{}\lq{}speed-up" to general purpose GPU programming. Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used - that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper - and we also made another reference timing by running our GPU implementations using a single thread. For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms. With respect to these reference timings, the speed-up is roughly 20,500x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.

>>>>

\begin{document}
\title{Acceleration of bilateral filtering algorithm for manycore and multicore architectures}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations. For multicore architectures, we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads.  We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions. We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency. We present empirical data on the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was (i) 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280 GPU and (ii) up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998} on multicore and manycore architectures. Bilateral filtering kernel is used to smooth an image without blurring its edges. It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003}, etc.

@@

In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time in half.
Our specific contributions are
\begin{enumerate}
\item
\item
\item
\end{enumerate}

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set neighbors per Center Pixel. A formal expression is given below:
\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

@@

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. Thus, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

@@

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined.

@@

It can be seen that after taking the $exp$ instruction out of the computation, the total sequential execution time is consistently less than 2\% of the original execution time with $exp$ instruction in place. For the smaller image it is strictly less than 1\% of the original time with $exp$ instruction in place. To be precise, it is 1.39\% maximum on Intel Nehalem for the larger image and 0.65\% for the smaller image on Core i7 processor. The lower memory transfer time for smaller image is due to the fact that the smaller image fits into the cache memory entirely and hence there are no repeated data transfers between the processor and the main memory.

@@

We have implemented a number of variations of bilateral filtering kernel for multicore chips, results of which are reported in section \ref{sec:results}. These algorithms are labeled as follows: 
\begin{itemize}
\item $BL\_Naive$ is the sequential version of the naive bilateral filtering algorithm, as described by Tomasi\cite{Tomasi1998}. 
\item $BL\_Tiles$ is an improved version of bilateral filtering kernel that employs data blocking to leverage the availability of multi-level cache hierarchy in the modern multicore CPUs. 
\item $BL\_Assembly$ is the version of naive bilateral filtering kernel implemented using assembly level code. 
\item $BL\_A\_Reduction$ is the version of bilateral filtering kernel that uses assembly language as well as reduction optimizations discussed in Section~\ref{sub:reduction}.
\item $BL\_Intrinsic$ version is an extension of $BL\_Assembly$ version where SSE intrinsics have replaced assembly level instructions.
\item $BL\_I\_Reduction$ is the extended version of $BL\_Intrinsic$ that employs intrinsics for reduction methods. These intrinsics can compute the sum of two 128 bit operands in parallel by exploiting the vector array processing. 
\end{itemize}
$BL\_Naive$ has a redundant calculation inside the innermost loop which has been moved out of loop for all other versions. The original instruction involving this calculation looks like
 $$gaussian\_ph = exp\left( \frac{-pd} {2 * sigma\_ph*sigma\_ph}\right);$$
After moving the redundant calculation out of the loop the same instruction becomes
 $$gaussian\_ph = exp( pd * pre\_calc\_sigma\_sq );$$

\subsubsection{Input Image}
In order to analyze the behavior of our kernel with varying image sizes, we have executed our kernel with two different images. One that fits into cache easily and the other that does not fit into cache but fits into main memory. Both of the images were converted to single spectrum images represented by single precision numbers for each pixel. The smaller image was 512 pixels wide and 512 pixels high, thus requiring 1MB memory to store it. As shown in Table~\ref{tab:archs} all of the multicore systems have cache memories bigger than 1 MB. The second image was 3000 pixels wide and 3000 pixels high and hence requires little more than 34 MB of memory to store it. None of the multicore cache memories were big enough to store this image entirely.
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Results}
\label{sec:results}
\subsection{Performance of bilateral filtering kernel on CPU}
For all of our experiments the filter radius was kept at 10, subsequently making the filter Window of size 21 x 21. The spatial spread and the photometric spread were also set to 10. The optimization flag $-O3$ was used for all experiments in addition to SSE2 for SIMDization. For Core i7 chip compiler flag SSE4.2 was used instead of SSE2. We do not use hyperthreading, i.e., there are no more than one thread per core.

@@

Moreover, when replacing shared memory transactions with dummy register transactions in our CUDA implementation of the pair-symmetric algorithm, we observed a consistent 40\% reduction in execution time. From these findings, we conclude that the inferior speed of GPU shared memory transactions to register transactions is what limited our pair-symmetric algorithm's performance gains to  a maximum of 33\%, i.e., reducing timings by no more than than 25\%.

To further verify this inference, we meticulously stepped through the Parallel Thread Execution (PTX) code, CUDA's quasi-assembly language, of our CUDA implementations to ensure that no unintended optimizations were being made by the compiler when substituting shared memory transactions with dummy register transactions. Not only were no optimizations being made but the PTX code showed that the shared memory transactions required additional load and store instructions, further evidencing the cost of using shared memory.

\textbf{GPU Speedup:} Another challenge in interpreting our results was porting the concept of \lq{}\lq{}speed-up" to general purpose GPU programming. Since speed-up is necessarily with respect to a reference timing, we used the same reference timing that our CPU optimizations used - that is how we calculated the 235.5x speed-up figure stated in the abstract of this paper - and we also made another reference timing by running our GPU implementations using a single thread. For example, using a single thread operating with filter width of 10 pixels to process a 768 x 512 pixel image, the time taken for the GTX280 to complete this execution was an average of 177.9 seconds for both of our algorithms; using a filter width of 5 pixels, the average execution time was 50.5 seconds for both algorithms. With respect to these reference timings, the speed-up is roughly 20,500x for the 10 pixel filter width and 12,440x for the 5 pixel filter width.
